Well-quasi-orderings on word languages

### 2025-10-15 src/amalgam.tex
Shorten section 6

\AP Our second focus is of practical nature: we want to give a decision
procedure for being \kl{well-quasi-ordered}. This will require us to introduce
\emph{effectiveness assumptions} on the \kl{amalgamation systems}. While most
of them will be innocuous, an important consequence is that we have to consider
\emph{classes of languages} rather than individual ones, for instance: the
class of all regular language, or the class of all context-free languages. Such
classes will be called \kl{effective amalgamative classes} (\kcref{effective
amalgamative classes}). In the following theorem, we prove that under such
assumptions, testing \kl{well-quasi-ordering} is inter-reducible to testing
whether a language of the class is empty, which is usually the simplest
problem for a computational model.

@@

\AP We say that a \kl{strongly effective amalgamative class} is an
\kl{effective amalgamative class} for which the emptiness problem is decidable.
Let us immediately remark that the class of \kl{regular languages} is
\kl(amalg){strongly effective}, and so is the class of context-free languages.
Therefore, \cref{infix-wqo-is-emptiness:thm} provides a concrete decision
procedure for these classes.

@@

\begin{definition}
    An \intro{amalgamation system}
    is a tuple $(\Sigma, R, \canrun, E)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    $\canrun \colon R \to (\Sigma \uplus \set{\cansep})^*$ is a 
    function computing a \intro{canonical decomposition} of a run,
    and $E$ describes the so-called \intro{admissible embeddings} between runs: If $\rho$ and $\sigma$ are runs from $R$, then $E(\rho, \sigma)$ is a subset of the subword embeddings between $\canrun[\rho]$ and $\canrun[\sigma]$. We write $\rho \intro*\runleq \sigma$ if $E(\rho, \sigma)$ is non-empty. If we want to refer to a specific embedding $f \in E(\rho, \sigma)$, we also write $\rho \runleq_f \sigma$.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, 
    the \intro{gap language} of $r$ at position $i$ is $\intro*\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item \emph{$(R, E)$ Forms a Category.}
            For all $\rho, \sigma, \tau \in R$,
            $\mathrm{id} \in E(\rho,\rho)$,
            and whenever $f \in E(\rho,\sigma)$ and $g \in E(\sigma,\tau)$,
            then $g \circ f \in E(\rho,\tau)$.
        \item \emph{Well-Quasi-Ordered System.}
            $(R, \runleq)$ is a well-quasi-ordered set.
        \item \emph{Concatenative Amalgamation.}
            Let $\rho_0, \rho_1, \rho_2$ be runs 
            with $\rho_0 \runleq_f \rho_1$ 
            and $\rho_0 \runleq_g \rho_2$.
            Then for all $0 \leq i \leq \card{\canrun(\rho_0)}$,
            there exists a run $\rho_3 \in R$ 
            and embeddings $\rho_1 \runleq_{g'} \rho_3$
            and $\rho_2 \runleq_{f'} \rho_3$ 
            satisfying two conditions:
            (a) $g' \circ f = f' \circ g$ (we write $h$ for this composition) and
            (b) for every $0 \leq j \leq \card{\canrun[\rho_0]}$, 
            the gap word $\GapWord{h}{j}$
            is either $\GapWord{f}{j} \GapWord{g}{j}$
            or $\GapWord{h}{j} = \GapWord{g}{j} \GapWord{f}{j}$. 
            Specifically, for $i$ we may fix $\GapWord{h}{i} = \GapWord{f}{i} \GapWord{g}{i}$.
            We refer to \cref{amalgamation-runs:fig} for an illustration 
            of this property.
    \end{enumerate}

@@

Intuitively, the definition of an amalgamation system allows the comparison of
runs, and the proper ``gluing'' of runs together to obtain new runs. A number
of well-known language classes can be seen to be recognized by \kl{amalgamation
systems}, e.g., regular languages \cite[Theorem 5.3]{ASZZ24}, reachability and
coverability languages of VASS \cite[Theorem 5.5]{ASZZ24}, and context-free
languages \cite[Theorem 5.10]{ASZZ24}. For this paper to be self-contained, we
will also recall how runs of a finite state automaton can be understood as an
\kl{amalgamation system}.

\begin{example}[{\cite[Section 3.2]{ASZZ24}}]
    Let $A = (Q, \delta, q_0, F)$ be a finite state automaton over a finite
    alphabet $\Sigma$. Let $\Delta$ be the set of transitions $(q_1, a, q_2)
    \in Q \times \Sigma \times Q$,
    and $R \subseteq \Delta^*$ be the set of 
    words over transitions that start with the initial state $q_0$,
    end in a final state $q_f \in F$, and such that the end state of a
    letter is the start state of the following one.
    The canonical decomposition $\canrun$
    is defined as a morphism from $\Delta^*$ to $\Sigma^*$
    that maps $(q,a,p)$ to $a$. 
    Because of the one-to-one correspondence of steps of a run $\rho$ and letters in its \kl{canonical decomposition}, 
    we may treat the two interchangeably.
    Finally, given two runs $\rho$ and $\sigma$ of the automaton,
    we say that an embedding $f \in \HigEmb(\canrun(\rho), \canrun(\sigma))$
    belongs to $E(\rho,\sigma)$ when
    $f$ is also defining an embedding from $\rho$ to $\sigma$ as words in $\Delta^*$.

    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
    whose language is precisely the language of words recognized
    by the automaton $A$.
\end{example}
\begin{proof}
    By definition, the embeddings inside $E(\rho,\sigma)$ are in
    of $\HigEmb(\canrun(\rho), \canrun(\sigma))$, and they compose properly.
    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
    according to Higman’s Lemma \cite{HIG52}.

    Let us now move to proving that the system satisfies the amalgamation
    property. Given three runs $\rho,\sigma,\tau \in R$, and two embeddings $f \in E(\rho,\sigma)$
    and $g \in E(\rho,\tau)$, we want to construct an amalgamated run $\sigma \vee \tau$.
    Because letters in the run $\rho$ respect the transitions of the automaton
    (i.e., if the letter $i$ ends in state $q$, then the letter $i+1$ starts in
    state $q$), then the \kl{gap word} at position $i$ starts in state $q$ and
    ends in state $q$ too. This means that for both embeddings
    $f$ and $g$, the \kl{gap words} are read by the automaton by looping
    on a state. In particular, these loops can be taken in any order
    and continue to represent a valid run. That is, we can even select
    the order of concatenation in the amalgamation for \emph{all} 
    $0 \leq i \leq \card{\canrun(\rho)}$ and not just for one separately.

    We conclude by remarking that 
    the language of this amalgamation system is
    the set of $\yieldrun(R)$, 
    because $R$ is the set of valid runs of the automaton,
    and $\yieldrun(\rho)$ is the word read along a run $\rho$.
\end{proof}

@@

	From \cref{pumping-periods:lem}, we may conclude that there are a finite number of words $x, y,$ and $w$ 
	such that each $w_k$ is contained in a language 
	$\InfPeriodChain{x}w\InfPeriodChain{y}$.

@@

\AP Let us now introduce our effectiveness assumptions on \kl{amalgamation
systems}. We follow the approach of \cite{ASZZ24} and require that an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$ is effective when $R$ is
recursively enumerable, the function $\canrun(\cdot)$ is computable, and for
any two runs $\rho, \sigma \in R$, the set $E(\rho,\sigma)$ is computable.

\AP We say that a class $\mathcal{C}$ of languages is an \intro{effective
amalgamative class} whenever for every $L \in \mathcal{C}$, there exists an
effective \kl{amalgamation system} recognizing $L$, and such that $\mathcal{C}$
is \kl{effectively closed under rational transductions}. Recall that a
\intro{rational transduction} is a \emph{relation} $R \subseteq \Sigma^* \times
\Gamma^*$ that can be defined by a (nondeterministic) finite automaton with
output \cite[Chapter 5, page 64]{BERST79}. A class of languages $\mathcal{C}$
is \intro{effectively closed under rational transductions} when, for every
language $L \in \mathcal{C}$, and every rational transduction $R \subseteq
\Sigma^* \times \Gamma^*$, the image of $L$ through $R$ --- that is, $\setof{v
\in \Gamma^*}{\exists u \in L. (u,v) \in R}$ --- is in $\mathcal{C}$ and
effectively computable. In particular, it implies that the class contains
languages over arbitrary (finite) alphabets.

\AP A class $\mathcal{C}$ of languages is \intro(amalg){strongly effective}
whenever the emptiness problem for languages in $\mathcal{C}$ is decidable.
This notion is interesting because usual language problems such as boundedness
or simultaneous boundedness are decidable for \kl{strongly effective
amalgamative classes}~\cite{ASZZ24}.

\begin{proofof}{infix-wqo-is-emptiness:thm}
	\cref{emptiness-decidable} We first show $\Rightarrow$ \cref{wqo-infix-decidable}. We aim to make the inclusion test of \cref{bounded-language:eq} of \cref{bounded-language:thm} effective. 
	Let $R(n,m,N_0) \defined \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in
    \Sigma^{\leq m \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup
    \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of
    the bounds $n$, $m$, and $N_0$, this language is regular. The map $L
    \mapsto L \cap \Sigma^* \setminus R(n,m,N_0)$  is a \kl{rational
    transduction} because $\Sigma^* \setminus R(n,m,N_0)$ is regular. Since
    $\mathcal{C}$ is \kl{closed under rational transductions}, we can therefore
    reduce the inclusion to emptiness of this language. However, we need to
    find these bounds first.

@@

Let us conclude by noting that it is unsurprisingly not possible to decide
whether a decidable language is \kl{well-quasi-ordered} by the \kl{prefix
relation}. This is a very easy result whose sole purpose is to contrast with
the decidability result of \cref{aut-cfg-infix:cor}.

\begin{remark}
    The following problem is undecidable: given a language $L$
    decided by a Turing machine, answer whether 
    $L$ is \kl{well-quasi-ordered} for the \kl{prefix relation}.
\end{remark}
\begin{proof}
    We reduce the halting problem on the empty string $\varepsilon$.
    Let $M$ be a Turing Machine, we write the languages $L$ of finite runs
    of $M$ starting on the empty string,
    that we surround by special markers. This language is decidable,
    and 
    is
    \kl{well-quasi-ordered} if and only if it is finite
    if and only if $M$ terminates on $\varepsilon$.
\end{proof}

>>>>

\AP Our second focus is of practical nature: we want to give a decision
procedure for being \kl{well-quasi-ordered}. This will require us to introduce
\emph{effectiveness assumptions} on the \kl{amalgamation systems}. While most
of them will be innocuous, an important consequence is that we have to consider
\emph{classes of languages} rather than individual ones, for instance: the
class of all regular language, or the class of all context-free languages. Such
classes will be called \kl[effectiveness assumptions@amalgamation]{effective amalgamative classes} (\kcref{effectiveness assumptions@amalgamation})\todo{can't get the knowledge to work}. In the following theorem, we prove that under such
assumptions, testing \kl{well-quasi-ordering} is inter-reducible to testing
whether a language of the class is empty, which is usually the simplest
problem for a computational model.

@@

%\AP We say that a \kl{strongly effective amalgamative class} is an
%\kl{effective amalgamative class} for which the emptiness problem is decidable.
%Let us immediately remark that the class of \kl{regular languages} is
%\kl(amalg){strongly effective}, and so is the class of context-free languages.
%Therefore, \cref{infix-wqo-is-emptiness:thm} provides a concrete decision
%procedure for these classes.

@@

\begin{definition}
    An \intro{amalgamation system}
    is a tuple $(\Sigma, R, \canrun, E)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    $\canrun \colon R \to (\Sigma \uplus \set{\cansep})^*$ is a 
    function computing a \intro{canonical decomposition} of a run,
    and $E$ describes the so-called \intro{admissible embeddings} between runs: If $\rho$ and $\sigma$ are runs from $R$, then $E(\rho, \sigma)$ is a subset of the subword embeddings between $\canrun(\rho)$ and $\canrun(\sigma)$. We write $\rho \intro*\runleq \sigma$ if $E(\rho, \sigma)$ is non-empty. If we want to refer to a specific embedding $f \in E(\rho, \sigma)$, we also write $\rho \runleq_f \sigma$.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, 
    the \intro{gap language} of $r$ at position $i$ is $\intro*\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item \emph{$(R, E)$ Forms a Category.}
            For all $\rho, \sigma, \tau \in R$,
            $\mathrm{id} \in E(\rho,\rho)$,
            and whenever $f \in E(\rho,\sigma)$ and $g \in E(\sigma,\tau)$,
            then $g \circ f \in E(\rho,\tau)$.
        \item \emph{Well-Quasi-Ordered System.}
            $(R, \runleq)$ is a well-quasi-ordered set.
        \item \emph{Concatenative Amalgamation.}
            Let $\rho_0, \rho_1, \rho_2$ be runs 
            with $\rho_0 \runleq_f \rho_1$ 
            and $\rho_0 \runleq_g \rho_2$.
            Then for all $0 \leq i \leq \card{\canrun(\rho_0)}$,
            there exists a run $\rho_3 \in R$ 
            and embeddings $\rho_1 \runleq_{g'} \rho_3$
            and $\rho_2 \runleq_{f'} \rho_3$ 
            satisfying two conditions:
            (a) $g' \circ f = f' \circ g$ (we write $h$ for this composition) and
            (b) for every $0 \leq j \leq \card{\canrun[\rho_0]}$, 
            the gap word $\GapWord{h}{j}$
            is either $\GapWord{f}{j} \GapWord{g}{j}$
            or $\GapWord{h}{j} = \GapWord{g}{j} \GapWord{f}{j}$. 
            Specifically, for $i$ we may fix $\GapWord{h}{i} = \GapWord{f}{i} \GapWord{g}{i}$.
            We refer to \cref{amalgamation-runs:fig} for an illustration 
            of this property.
    \end{enumerate}

@@

Intuitively, the definition of an amalgamation system allows the comparison of
runs, and the proper ``gluing'' of runs together to obtain new runs. A number
of well-known language classes can be seen to be recognized by \kl{amalgamation
systems}, e.g., regular languages \cite[Theorem 5.3]{ASZZ24}, reachability and
coverability languages of VASS \cite[Theorem 5.5]{ASZZ24}, and context-free
languages \cite[Theorem 5.10]{ASZZ24}.

%For this paper to be self-contained, we
%will also recall how runs of a finite state automaton can be understood as an
%\kl{amalgamation system}.
%
%\begin{example}[{\cite[Section 3.2]{ASZZ24}}]
%    Let $A = (Q, \delta, q_0, F)$ be a finite state automaton over a finite
%    alphabet $\Sigma$. Let $\Delta$ be the set of transitions $(q_1, a, q_2)
%    \in Q \times \Sigma \times Q$,
%    and $R \subseteq \Delta^*$ be the set of 
%    words over transitions that start with the initial state $q_0$,
%    end in a final state $q_f \in F$, and such that the end state of a
%    letter is the start state of the following one.
%    The canonical decomposition $\canrun$
%    is defined as a morphism from $\Delta^*$ to $\Sigma^*$
%    that maps $(q,a,p)$ to $a$. 
%    Because of the one-to-one correspondence of steps of a run $\rho$ and letters in its \kl{canonical decomposition}, 
%    we may treat the two interchangeably.
%    Finally, given two runs $\rho$ and $\sigma$ of the automaton,
%    we say that an embedding $f \in \HigEmb(\canrun(\rho), \canrun(\sigma))$
%    belongs to $E(\rho,\sigma)$ when
%    $f$ is also defining an embedding from $\rho$ to $\sigma$ as words in $\Delta^*$.
%
%    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
%    whose language is precisely the language of words recognized
%    by the automaton $A$.
%\end{example}
%\begin{proof}
%    By definition, the embeddings inside $E(\rho,\sigma)$ are in
%    of $\HigEmb(\canrun(\rho), \canrun(\sigma))$, and they compose properly.
%    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
%    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
%    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
%    according to Higman’s Lemma \cite{HIG52}.
%    
%    Let us now move to proving that the system satisfies the amalgamation
%    property. Given three runs $\rho,\sigma,\tau \in R$, and two embeddings $f \in E(\rho,\sigma)$
%    and $g \in E(\rho,\tau)$, we want to construct an amalgamated run $\sigma \vee \tau$.
%    Because letters in the run $\rho$ respect the transitions of the automaton
%    (i.e., if the letter $i$ ends in state $q$, then the letter $i+1$ starts in
%    state $q$), then the \kl{gap word} at position $i$ starts in state $q$ and
%    ends in state $q$ too. This means that for both embeddings
%    $f$ and $g$, the \kl{gap words} are read by the automaton by looping
%    on a state. In particular, these loops can be taken in any order
%    and continue to represent a valid run. That is, we can even select
%    the order of concatenation in the amalgamation for \emph{all} 
%    $0 \leq i \leq \card{\canrun(\rho)}$ and not just for one separately.
%
%    We conclude by remarking that 
%    the language of this amalgamation system is
%    the set of $\yieldrun(R)$, 
%    because $R$ is the set of valid runs of the automaton,
%    and $\yieldrun(\rho)$ is the word read along a run $\rho$.
%\end{proof}

@@

	\todo[inline]{The following can probably be replaced by a reference to Fine \& Wilf}
	From \cref{pumping-periods:lem}, we may conclude that there are a finite number of words $x, y,$ and $w$ 
	such that each $w_k$ is contained in a language 
	$\InfPeriodChain{x}w\InfPeriodChain{y}$.

@@

\AP To construct a decision procedure for well-quasi-orderedness under $\infleq$, we 
our \kl{amalgamation systems} to satisfy certain \intro{effectiveness assumptions}. We require that for an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$, $R$ is
recursively enumerable, the function $\canrun(\cdot)$ is computable, and for
any two runs $\rho, \sigma \in R$, the set $E(\rho,\sigma)$ is computable. Additionally, we require the class to be effectively closed under rational transductions \cite[Chapter 5, page 64]{BERST79}.

%\AP We say that a class $\mathcal{C}$ of languages is an \intro{effective
%amalgamative class} whenever for every $L \in \mathcal{C}$, there exists an
%effective \kl{amalgamation system} recognizing $L$, and such that $\mathcal{C}$
%is \kl{effectively closed under rational transductions}. Recall that a
%\intro{rational transduction} is a \emph{relation} $R \subseteq \Sigma^* \times
%\Gamma^*$ that can be defined by a (nondeterministic) finite automaton with
%output \cite[Chapter 5, page 64]{BERST79}. A class of languages $\mathcal{C}$
%is \intro{effectively closed under rational transductions} when, for every
%language $L \in \mathcal{C}$, and every rational transduction $R \subseteq
%\Sigma^* \times \Gamma^*$, the image of $L$ through $R$ --- that is, $\setof{v
%\in \Gamma^*}{\exists u \in L. (u,v) \in R}$ --- is in $\mathcal{C}$ and
%effectively computable. In particular, it implies that the class contains
%languages over arbitrary (finite) alphabets.

%\AP A class $\mathcal{C}$ of languages is \intro(amalg){strongly effective}
%whenever the emptiness problem for languages in $\mathcal{C}$ is decidable.
%This notion is interesting because usual language problems such as boundedness
%or simultaneous boundedness are decidable for \kl{strongly effective
%amalgamative classes}~\cite{ASZZ24}.

\begin{proofof}{infix-wqo-is-emptiness:thm}
	We first show \cref{emptiness-decidable} $\Rightarrow$ \cref{wqo-infix-decidable}. We aim to make the inclusion test of \cref{bounded-language:eq} of \cref{bounded-language:thm} effective. 
	Let $R(n,m,N_0) \defined \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in
    \Sigma^{\leq m \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup
    \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of
    the bounds $n$, $m$, and $N_0$, this language is regular. The map $L
    \mapsto L \cap \Sigma^* \setminus R(n,m,N_0)$  is a \kl{rational
    transduction} because $\Sigma^* \setminus R(n,m,N_0)$ is regular. Since
    $\mathcal{C}$ is \kl{closed under rational transductions}, we can therefore
    reduce the inclusion to emptiness of this language. However, we need to
    find these bounds first.

@@

\begin{remark}
	Let us conclude by noting that it is unsurprisingly not possible to decide
	whether a decidable language is \kl{well-quasi-ordered} by the \kl{prefix
	relation} in general. 
	%This is a very easy result whose sole purpose is to contrast with
	%the decidability result of \cref{aut-cfg-infix:cor}.

\end{remark}
\begin{proof}
    We reduce the halting problem on the empty string $\varepsilon$.
    Let $M$ be a Turing Machine, we write the languages $L$ of finite runs
    of $M$ starting on the empty string,
    that we surround by special markers. This language is decidable,
    and 
    is
    \kl{well-quasi-ordered} if and only if it is finite
    if and only if $M$ terminates on $\varepsilon$.
\end{proof}
